---
title: "HOBO Logger Data"
author: "Brett D. Jameson"
date: "2024-08-29"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(tidyverse)
library(knitr)
library(plyr)
```

## Mangrove Biogeochemistry (MBGC) Project - Microprofile Data Processing and Visualization

This R Markdown file contains code scripts associated with the processing of HOBO temperature logger data. 

Data were collected as part of the 2024 Mangrove Biogeochemistry project. Freshly collected mangrove sediment cores were incubated daily under varying nutrient amendment regimes and profiled to characterize vertical distributions of N2O, NO, H2S, and O2. Porewater profiles were obtained using Clark-type microelectrodes from Unisense.

# Data input
```{r}
path.p = "./hobo-data/" #the location of HOBO files for all mangrove experiments

# List data files
file.names <- list.files(path = path.p, pattern = "csv$")  # list all csv file names in the folder
#file.names <- file.names[!grepl("metadata", file.names)]   # omit metadata from files to be read in as data

newnames <- c("obs", "date-time", "tempC", "coupler-attached", "host-connected", "stopped", "end-of-file") # create column names

df <- tibble(file.name = file.names) %>%
  mutate(experiment = gsub("^([^_]*_[^_]*)_.*$", "\\1", file.name)) %>% # Get experiment ID from filename
  mutate(flume = gsub("^[^_]+_[^_]+_[^_]+_[^_]+_([^_]+)_.*", "\\1", file.name)) %>% # Get flume ID from filename
  mutate(data = map(file.name, ~read_csv(file.path(path.p, .), col_names = newnames, skip=2)))   # Get associated temp log data
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
