---
title: "Microprofile data"
author: "Brett D. Jameson"
date: "2024-06-28"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(tidyverse)
library(knitr)
library(plyr)
library(plotrix)
```

## Mangrove Biogeochemistry (MBGC) Project - Microprofile Data Processing and Visualization

This R Markdown file contains code scripts associated with the processing of sediment microprofile data. 

Data were collected as part of the 2024 Mangrove Biogeochemistry project. Freshly collected mangrove sediment cores were incubated daily under varying nutrient amendment regimes and profiled to characterize vertical distributions of N2O, NO, H2S, and O2. Porewater profiles were obtained using Clark-type microelectrodes from Unisense.

# Data input
```{r Data input}
# Read compiled microprofile data
profiles <- read.csv(file = "./compiled-data/MBGC.mangrove.microprofiles.2024.csv")
profiles <- profiles %>% subset(experiment != "Mangrove_1")
```

## Data parsing

Subset data frame by analyte.

```{r pressure, echo=FALSE}
sulfide <- subset(profiles, analyte == "H2S")
nitrous <- subset(profiles, analyte == "N2O")
oxygen <- subset(profiles, analyte == "O2")
nitric <- subset(profiles, analyte == "NO")
```

## Exploratory data plots (raw data)

Oxygen profiles.

```{r}
oxygen %>% #subset(experiment == "Mangrove_3" & light.dark == "Dark") %>%
  ggplot(aes(mV, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(4,-1)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  #scale_x_continuous(limits = c(5,9)) +
  facet_wrap(~experiment + treatment) +
  theme_bw()
```

Nitrous oxide profiles.

```{r}
nitrous %>% #subset(experiment == "Mangrove_2" & light.dark == "Dark") %>%
  ggplot(aes(mV, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(10,-1)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(3,12)) +
  facet_wrap(~experiment + treatment) +
  theme_bw()
```

Nitric oxide profiles.

```{r}
nitric %>% #subset(experiment == "Mangrove_6" & light.dark == "Light" & channel == 1) %>%
  ggplot(aes(mV, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(8,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(1,20)) +
  facet_wrap(~experiment + treatment) +
  theme_bw()

nitric %>% subset(experiment == "Mangrove_6" & light.dark == "Dark" & channel == 1) %>%
  ggplot(aes(mV, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(8,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(3,20)) +
  facet_wrap(~treatment) +
  theme_bw()
```

Sulfide Profiles.

```{r}
 sulfide %>% subset(depth > 15000) %>%
  ggplot(aes(mV, depth, color=as.factor(core))) +
  geom_point() +
  scale_y_reverse() +
  facet_wrap(~treatment)
```
# Apply calibrations to raw sensor data

Read and subset calibration data.
```{r}
cal.data <- read.csv(file = "./compiled-data/MBGC.sensor.calibrations.2024.csv")

nitrous.cal <- cal.data %>% subset(analyte == "N2O")
ox.cal <- cal.data %>% subset(analyte == "O2")
nitric.cal <- cal.data %>% subset(analyte == "NO")
sulf.cal <- cal.data %>% subset(analyte == "H2S")
```

# Calibration data plots 

Plot oxygen calibration data. ERROR BARS ON THESE PLOTS LOOK SUS
```{r}
# Calculate mean +/- SD for calibration data 
ox.cal.sum <- ox.cal %>%
  plyr::ddply(.(experiment, channel, light.dark, cal.ID, concentration), summarize,
                    mean.mV = mean(signal.mV, na.rm = TRUE), stdev.mV=sd(signal.mV, na.rm = TRUE))

ox.cal.sum %>% #subset(experiment == "Mangrove_2") %>%
  ggplot(aes(concentration, mean.mV, color = as.factor(cal.ID), shape = as.factor(channel))) +
    geom_errorbar(aes(ymin = mean.mV - stdev.mV, ymax = mean.mV + stdev.mV, x = concentration), height = 0.25) +
    geom_point(size = 2.5) +
    geom_line() +
    scale_shape_manual(values = c(0,1,2,3)) +
    facet_wrap(~light.dark + experiment)
```

Plot NO calibration data.
```{r}
nitric.cal.sum <- nitric.cal %>%
  plyr::ddply(.(experiment, channel, light.dark, cal.ID, concentration), summarize,
                    mean.mV = mean(signal.mV, na.rm = TRUE), stdev.mV=sd(signal.mV, na.rm = TRUE))

nitric.cal.sum %>% #subset(experiment == "Mangrove_6") %>%
  ggplot(aes(concentration, mean.mV, color = as.factor(cal.ID), shape = as.factor(channel))) +
    geom_errorbar(aes(ymin = mean.mV - stdev.mV, ymax = mean.mV + stdev.mV)) +
    geom_point(size = 2.5) +
    geom_line() +
    scale_shape_manual(values = c(0,1,2,3)) +
    facet_wrap(~light.dark + experiment)
```

Plot N2O calibration data.
```{r}
nitrous.cal.sum <- nitrous.cal %>%
  plyr::ddply(.(experiment, channel, light.dark, cal.ID, concentration), summarize,
                    mean.mV = mean(signal.mV, na.rm = TRUE), stdev.mV=sd(signal.mV, na.rm = TRUE))

nitrous.cal.sum %>% #subset(experiment == "Mangrove_2") %>%
  ggplot(aes(concentration, mean.mV, color = as.factor(cal.ID), shape = as.factor(channel))) +
    geom_errorbar(aes(ymin = mean.mV - stdev.mV, ymax = mean.mV + stdev.mV)) +
    geom_point(size = 2.5) +
    geom_line() +
    scale_shape_manual(values = c(0,1,2,3)) +
    facet_wrap(~light.dark + experiment)
```

Plot H2S calibration data.
```{r}
sulf.cal.sum <- sulf.cal %>%
  plyr::ddply(.(experiment, channel, light.dark, cal.ID, concentration), summarize,
                    mean.mV = mean(signal.mV, na.rm = TRUE), stdev.mV=sd(signal.mV, na.rm = TRUE))

sulf.cal.sum %>% #subset(experiment == "Mangrove_2") %>%
  ggplot(aes(concentration, mean.mV, color = as.factor(cal.ID), shape = as.factor(channel))) +
    geom_errorbar(aes(ymin = mean.mV - stdev.mV, ymax = mean.mV + stdev.mV)) +
    geom_point(size = 2.5) +
    geom_line() +
    scale_shape_manual(values = c(0,1,2,3)) +
    facet_wrap(~light.dark + experiment)
```

# Apply calibrations to raw data

Fit linear models to calibration data to obtain coefficients, then merge calibration coefficients with raw sensor data. Finally, apply calibration coefficients to calculate concentration profiles.

## 1. Oxygen profiles

```{r}
ox_models <- ox.cal.sum %>% 
  group_by(experiment, channel, light.dark, cal.ID) %>% 
  do(data.frame(., as.list(coef(lm(mean.mV~concentration, data = .))))) %>%
  rename_at(8:9, ~c("intercept", "slope"))

oxygen <- ox_models %>% subset(concentration == 0) %>% select(-mean.mV, -stdev.mV) %>%
  group_by(experiment, channel, light.dark) %>%
  pivot_wider(names_from = cal.ID, values_from = c(intercept, slope), names_vary = "slowest") %>%
  right_join(., oxygen, by = join_by(experiment, channel, light.dark)) %>%
  #subset(experiment == 'Mangrove_6') %>%
  select(experiment, channel, light.dark, incubation, treatment, 
         flume, core, replicate, analyte, time, depth, mV,
         intercept_1, slope_1)

# calculate concentration for each calibration curve
oxygen$OX_1 <- (oxygen$mV - oxygen$intercept_1) / oxygen$slope_1
```

Plot calibrated Ox data.
```{r}
oxygen %>% subset(experiment == "Mangrove_2" & light.dark == "Light") %>%
  ggplot(aes(OX_1, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(4,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = expression(O[2]~"("*mu*"mol L"^-1*")")) +
  #scale_x_continuous(limits = c(3,20)) +
  facet_wrap(~treatment) +
  theme_bw()

oxygen %>% subset(experiment == "Mangrove_2" & light.dark == "Dark") %>%
  ggplot(aes(OX_1, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(4,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = expression(O[2]~"("*mu*"mol L"^-1*")")) +
  #scale_x_continuous(limits = c(0,20)) +
  facet_wrap(~treatment) +
  theme_bw()
```

## 2. NO Profiles

```{r}
nitric_models <- nitric.cal.sum %>% 
  group_by(experiment, channel, light.dark, cal.ID) %>% 
  do(data.frame(., as.list(coef(lm(mean.mV~concentration, data = .))))) %>%
  rename_at(8:9, ~c("intercept", "slope"))

nitric <- nitric_models %>% subset(concentration == 0) %>% select(-mean.mV, -stdev.mV) %>%
  group_by(experiment, channel, light.dark) %>%
  pivot_wider(names_from = cal.ID, values_from = c(intercept, slope), names_vary = "slowest") %>%
  right_join(., nitric, by = join_by(experiment, channel, light.dark)) %>%
  #subset(experiment == 'Mangrove_6') %>%
  select(experiment, channel, light.dark, incubation, treatment, 
         flume, core, replicate, analyte, time, depth, mV,
         intercept_1, slope_1, intercept_2, slope_2)

# calculate concentration for each calibration curve
nitric$NO_1 <- (nitric$mV - nitric$intercept_1) / nitric$slope_1
nitric$NO_2 <- (nitric$mV - nitric$intercept_2) / nitric$slope_2
```

Plot calibrated NO data.
```{r}
nitric %>% subset(experiment == "Mangrove_15" & light.dark == "Dark" & channel == 4) %>%
  ggplot(aes(NO_1, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(8,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = expression("NO (nmol"~L^-1*")")) +
  #scale_x_continuous(limits = c(3,20)) +
  facet_wrap(~experiment + treatment) +
  theme_bw()

nitric %>% subset(light.dark == "Dark") %>%
  ggplot(aes(NO_2, depth/1000, color=as.factor(core))) +
  geom_point(aes(shape = as.factor(channel))) +
  geom_path(aes(shape = as.factor(channel))) +
  scale_y_reverse(limits = c(8,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = expression("NO (nmol"~L^-1*")")) +
  scale_x_continuous(limits = c(0,3000)) +
  facet_wrap(~treatment+channel, ncol = 2) +
  theme_bw()
```

## 3. N2O Profiles

```{r}
nitrous_models <- nitrous.cal.sum %>% 
  group_by(experiment, channel, light.dark, cal.ID) %>% 
  do(data.frame(., as.list(coef(lm(mean.mV~concentration, data = .))))) %>%
  rename_at(8:9, ~c("intercept", "slope"))

nitrous <- nitrous_models %>% subset(concentration == 0) %>% select(-mean.mV, -stdev.mV) %>%
  group_by(experiment, channel, light.dark) %>%
  pivot_wider(names_from = cal.ID, values_from = c(intercept, slope), names_vary = "slowest") %>%
  right_join(., nitrous, by = join_by(experiment, channel, light.dark)) %>%
  subset(experiment != 'Mangrove_1') %>%
  select(experiment, channel, light.dark, incubation, treatment, 
         flume, core, replicate, analyte, time, depth, mV,
         intercept_1, slope_1)

# calculate concentration for each calibration curve
nitrous$N2O_1 <- (nitrous$mV - nitrous$intercept_1) / nitrous$slope_1

nitrous_subset <- nitrous %>% subset(experiment == "Mangrove_6" & experiment == "Mangrove_15")
```

Plot calibrated N2O data.
```{r}
nitrous %>% subset(experiment == "Mangrove_8" & light.dark == "Light") %>%
  ggplot(aes(N2O_1*1000, depth/1000, color=as.factor(core), shape=as.factor(channel))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(10,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(-15,100)) +
  facet_wrap(~treatment) +
  theme_bw()

nitrous %>% subset(experiment == "Mangrove_8" & light.dark == "Dark") %>%
  ggplot(aes(N2O_1*1000, depth/1000, color=as.factor(core), shape=as.factor(channel))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(10,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(-15,50)) +
  facet_wrap(~treatment) +
  theme_bw()

nitrous %>% subset(experiment == "Mangrove_14" & light.dark == "Dark") %>%
  ggplot(aes(N2O_1*1000, depth/1000, color=as.factor(core), shape=as.factor(channel))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(10,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(-15,70)) +
  facet_wrap(~treatment) +
  theme_bw()
```

## 3. H2S Profiles

```{r}
sulf_models <- sulf.cal.sum %>% 
  group_by(experiment, channel, light.dark, cal.ID) %>% 
  do(data.frame(., as.list(coef(lm(mean.mV~concentration, data = .))))) %>%
  rename_at(8:9, ~c("intercept", "slope"))

sulfide <- sulf_models %>% subset(concentration == 0) %>% select(-mean.mV, -stdev.mV) %>%
  group_by(experiment, channel, light.dark) %>%
  pivot_wider(names_from = cal.ID, values_from = c(intercept, slope), names_vary = "slowest") %>%
  right_join(., sulfide, by = join_by(experiment, channel, light.dark)) %>%
  subset(experiment != 'Mangrove_2') %>%
  select(experiment, channel, light.dark, incubation, treatment, 
         flume, core, replicate, analyte, time, depth, mV,
         intercept_1, slope_1, intercept_2, slope_2, intercept_3, slope_3)

# calculate concentration for each calibration curve
sulfide$H2S_1 <- (sulfide$mV - sulfide$intercept_1) / sulfide$slope_1
sulfide$H2S_2 <- (sulfide$mV - sulfide$intercept_2) / sulfide$slope_2
```

Plot calibrated N2O data.
```{r}
sulfide %>% subset(replicate == 1 & experiment == "Mangrove_11" | experiment == "Mangrove_12") %>%
  ggplot(aes(H2S_1, depth/1000, color=as.factor(core), shape=as.factor(channel))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(30,10)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(-5,200)) +
  facet_wrap(~experiment + flume) +
  theme_bw()
```


# Preliminary profiles for Tom

```{r}
nitric <- read.csv(file = "./compiled-data/Mangrove_6_15_NO_adjusted_revised.csv")
nitrous <- read.csv(file = "./compiled-data/Mangrove_8_14_N2O_adjusted.csv")
```

Pre-hurricane Nitric oxid
```{r}
nitric_pre <- nitric %>% subset(experiment == "Mangrove_6" & channel == 4)
nitric_1 <- nitric_pre %>% filter(flume == 1) 
nitric_2 <- nitric_pre %>% filter(flume == 2)
nitric_3 <- nitric_pre %>% filter(!row_number() %in% c(820:861) & flume == 3)
nitric_4 <- nitric_pre %>% filter(flume == 4)
nitric_pre <- rbind(nitric_1, nitric_2, nitric_3, nitric_4)
```

```{r}
nitric_pre_avg <- ddply(nitric_pre, .(experiment, treatment, depth, light.dark), summarize,
                    NO_mean = mean(NO_adj, na.rm = TRUE), NO_stdev=sd(NO_adj, na.rm = TRUE), 
                    NO_sterr=std.error(NO_adj))
```

Nitric oxide profiles
```{r}
nitric_pre %>% 
  ggplot(aes(NO_adj, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(10,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = expression("NO (nmol"~L^-1*")")) +
  scale_x_continuous(limits = c(0,10)) +
  facet_wrap(~treatment + light.dark, ncol = 2) +
  theme_bw()

nitric_pre_avg %>% 
  ggplot(aes(NO_mean, depth/1000, color = as.factor(treatment))) +
  geom_hline(yintercept = 0) +
  geom_errorbarh(aes(xmin = NO_mean - NO_sterr/1.5, xmax = NO_mean + NO_sterr/2))+
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(10,-2)) +
  scale_color_discrete(name = expression("[NO"[3]^-1*"]")) +
  labs(y = "Depth (mm)", x = expression("NO ("*mu*"mol"~L^-1*")")) +
  #scale_x_continuous(limits = c(3,20)) +
  facet_wrap(~treatment + light.dark, ncol = 2) +
  theme_bw() +
  theme(legend.background = element_rect(color = "black"),
        legend.position = c(0.90,0.075),
        legend.key.size = unit(1, "mm"))
```

Nitrous oxide
```{r}
nitrous_pre <- nitrous %>% subset(experiment == "Mangrove_8")
nitrous_pre <- nitrous_pre %>% filter(!row_number() %in% c(1:25) & !row_number() %in% c(638:663))
```

```{r}
nitrous_pre_avg <- ddply(nitrous_pre, .(experiment, treatment, depth, light.dark), summarize,
                    N2O_mean = mean(N2O_adj, na.rm = TRUE), N2O_stdev=sd(N2O_adj, na.rm = TRUE), 
                    N2O_sterr=std.error(N2O_adj))
nitrous_pre_avg <- nitrous_pre_avg %>% arrange(treatment, light.dark) 
  
#write.csv(nitrous_pre_avg, "MBGC.N2O.pre.hurricane.csv")
```

```{r}
nitrous_pre %>% 
  ggplot(aes(N2O_adj*1000, depth/1000, color=as.factor(core), shape = as.factor(channel))) +
  geom_hline(yintercept = 0) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(10,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = expression("NO (nmol"~L^-1*")")) +
  #scale_x_continuous(limits = c(0,10)) +
  facet_wrap(~treatment + light.dark, ncol = 2) +
  theme_bw()

nitrous_pre_avg %>% 
  ggplot(aes(N2O_mean*1000, depth/1000, color = as.factor(treatment))) +
  geom_hline(yintercept = 0) +
  geom_errorbarh(aes(xmin = N2O_mean*1000 - N2O_sterr*1000, xmax = N2O_mean*1000 + N2O_sterr*1000))+
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(10,-2)) +
  scale_color_discrete(name = expression("[NO"[3]^-1*"]")) +
  labs(y = "Depth (mm)", x = expression("N"[2]*"O (nmol"~L^-1*")")) +
  #scale_x_continuous(limits = c(3,20)) +
  facet_wrap(~treatment + light.dark, ncol = 2) +
  theme_bw() +
  theme(legend.background = element_rect(color = "black"),
        legend.position = c(0.90,0.075),
        legend.key.size = unit(1, "mm"))
```


```{r}
nitrous_trim <- nitrous_pre_avg %>% 
  subset(treatment == "50" | treatment == "200") %>%
  mutate(depth = depth / 1000)

nitric_trim <- nitric_pre_avg %>% 
  subset(treatment == "50" | treatment == "200") %>%
  mutate(NO_sterr = NO_sterr / 2, depth = depth / 1000)

combined <-
  
  ggplot() +
  
  geom_errorbarh(data = nitric_trim, 
                 aes(y = depth, xmin = NO_mean - NO_sterr, xmax = NO_mean + NO_sterr), 
                 color =  "darkgrey") +
  geom_point(data = nitric_trim, aes(NO_mean, depth), colour = "darkred", size=1) + 
  geom_path(data = nitric_trim, aes(NO_mean, depth), colour = "darkred", size=1) + 

  
  geom_errorbarh(data = nitrous_trim, 
                 aes(y = depth, xmin = N2O_mean*50 - N2O_sterr*50, xmax = N2O_mean*50 + N2O_sterr*50),
                 color = "darkgrey") +
  geom_point(data = nitrous_trim, aes(N2O_mean*50, depth), colour = "darkblue") + 
  geom_path(data = nitrous_trim, aes(N2O_mean*50, depth), colour = "darkblue") + 

  
  scale_y_reverse() +
  labs(y ="Depth (mm)", x = expression("NO ("*mu*"mol"~"L"^-1*")")) +
  scale_x_continuous(sec.axis=sec_axis(~ .*1/50, name=expression(N[2]*"O ("*mu*"mol"~"L"^-1*")"))) +
  facet_wrap(~treatment + light.dark, ncol = 2) +
  theme_bw() +
  theme(strip.background = element_blank(), strip.text.x = element_blank(),
        axis.title.x.top = element_text(color = "darkblue"),
        axis.title.x.bottom = element_text(color = "darkred"))
combined
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
