---
title: "Microprofile data"
author: "Brett D. Jameson"
date: "2024-06-28"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(tidyverse)
library(knitr)
library(plyr)
```

## Mangrove Biogeochemistry (MBGC) Project - Microprofile Data Processing and Visualization

This R Markdown file contains code scripts associated with the processing of sediment microprofile data. 

Data were collected as part of the 2024 Mangrove Biogeochemistry project. Freshly collected mangrove sediment cores were incubated daily under varying nutrient amendment regimes and profiled to characterize vertical distributions of N2O, NO, H2S, and O2. Porewater profiles were obtained using Clark-type microelectrodes from Unisense.

# Data input
```{r Data input}
# Read compiled microprofile data
profiles <- read.csv(file = "./compiled-data/MBGC.mangrove.microprofiles.2024.csv")
```

## Data parsing

Subset data frame by analyte.

```{r pressure, echo=FALSE}
sulfide <- subset(profiles, analyte == "Sulfide")
nitrous <- subset(profiles, analyte == "N2O")
oxygen <- subset(profiles, analyte == "O2")
nitric <- subset(profiles, analyte == "NO")
```

## Exploratory data plots (raw data)

Oxygen profiles.

```{r}
oxygen %>% subset(experiment == "Mangrove_3" & light.dark == "Dark") %>%
  ggplot(aes(mV, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(4,-1)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  #scale_x_continuous(limits = c(5,9)) +
  facet_wrap(~experiment + treatment) +
  theme_bw()
```

Nitrous oxide profiles.

```{r}
nitrous %>% subset(experiment == "Mangrove_2" & light.dark == "Dark") %>%
  ggplot(aes(mV, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(10,-1)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(5,9)) +
  facet_wrap(~treatment) +
  theme_bw()
```

Nitric oxide profiles.

```{r}
nitric %>% subset(experiment == "Mangrove_6" & light.dark == "Light" & channel == 1) %>%
  ggplot(aes(mV, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(8,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(3,20)) +
  facet_wrap(~treatment) +
  theme_bw()

nitric %>% subset(experiment == "Mangrove_6" & light.dark == "Dark" & channel == 1) %>%
  ggplot(aes(mV, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(8,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(3,20)) +
  facet_wrap(~treatment) +
  theme_bw()
```

Sulfide Profiles.

```{r}
 sulfide %>% subset(depth > 15000) %>%
  ggplot(aes(mV, depth, color=as.factor(core))) +
  geom_point() +
  scale_y_reverse() +
  facet_wrap(~treatment)
```
# Apply calibrations to raw sensor data

Read and subset calibration data.
```{r}
cal.data <- read.csv(file = "./compiled-data/MBGC.sensor.calibrations.2024.csv")

nitrous.cal <- cal.data %>% subset(analyte == "N2O")
ox.cal <- cal.data %>% subset(analyte == "O2")
nitric.cal <- cal.data %>% subset(analyte == "NO")
sulf.cal <- cal.data %>% subset(analyte == "Sulfide")
```

# Calibration data plots 

Plot oxygen calibration data. ERROR BARS ON THESE PLOTS LOOK SUS
```{r}
# Calculate mean +/- SD for calibration data 
ox.cal.sum <- ox.cal %>%
  plyr::ddply(.(experiment, channel, light.dark, cal.ID, concentration), summarize,
                    mean.mV = mean(signal.mV, na.rm = TRUE), stdev.mV=sd(signal.mV, na.rm = TRUE))

ox.cal.sum %>% #subset(experiment == "Mangrove_2") %>%
  ggplot(aes(concentration, mean.mV, color = as.factor(cal.ID), shape = as.factor(channel))) +
    geom_errorbar(aes(ymin = mean.mV - stdev.mV, ymax = mean.mV + stdev.mV, x = concentration), height = 0.25) +
    geom_point(size = 2.5) +
    geom_line() +
    scale_shape_manual(values = c(0,1,2,3)) +
    facet_wrap(~light.dark + experiment)
```

Plot NO calibration data.
```{r}
nitric.cal.sum <- nitric.cal %>%
  plyr::ddply(.(experiment, channel, light.dark, cal.ID, concentration), summarize,
                    mean.mV = mean(signal.mV, na.rm = TRUE), stdev.mV=sd(signal.mV, na.rm = TRUE))

nitric.cal.sum %>% subset(experiment == "Mangrove_6") %>%
  ggplot(aes(concentration, mean.mV, color = as.factor(cal.ID), shape = as.factor(channel))) +
    geom_errorbar(aes(ymin = mean.mV - stdev.mV, ymax = mean.mV + stdev.mV)) +
    geom_point(size = 2.5) +
    geom_line() +
    scale_shape_manual(values = c(0,1,2,3)) +
    facet_wrap(~light.dark)
```

Plot N2O calibration data.
```{r}
nitrous.cal.sum <- nitrous.cal %>%
  plyr::ddply(.(experiment, channel, light.dark, cal.ID, concentration), summarize,
                    mean.mV = mean(signal.mV, na.rm = TRUE), stdev.mV=sd(signal.mV, na.rm = TRUE))

nitrous.cal.sum %>% subset(experiment == "Mangrove_2") %>%
  ggplot(aes(concentration, mean.mV, color = as.factor(cal.ID), shape = as.factor(channel))) +
    geom_errorbar(aes(ymin = mean.mV - stdev.mV, ymax = mean.mV + stdev.mV)) +
    geom_point(size = 2.5) +
    geom_line() +
    scale_shape_manual(values = c(0,1,2,3)) +
    facet_wrap(~light.dark)
```

# Apply calibrations to raw data

Fit linear models to calibration data to obtain coefficients, then merge calibration coefficients with raw sensor data. Finally, apply calibration coefficients to calculate concentration profiles.

## 1. Oxygen profiles

```{r}
ox_models <- ox.cal.sum %>% 
  group_by(experiment, channel, light.dark, cal.ID) %>% 
  do(data.frame(., as.list(coef(lm(mean.mV~concentration, data = .))))) %>%
  rename_at(8:9, ~c("intercept", "slope"))

oxygen <- ox_models %>% subset(concentration == 0) %>% select(-mean.mV, -stdev.mV) %>%
  group_by(experiment, channel, light.dark) %>%
  pivot_wider(names_from = cal.ID, values_from = c(intercept, slope), names_vary = "slowest") %>%
  right_join(., oxygen, by = join_by(experiment, channel, light.dark)) %>%
  #subset(experiment == 'Mangrove_6') %>%
  select(experiment, channel, light.dark, incubation, treatment, 
         flume, core, replicate, analyte, time, depth, mV,
         intercept_1, slope_1)

# calculate concentration for each calibration curve
oxygen$OX_1 <- (oxygen$mV - oxygen$intercept_1) / oxygen$slope_1
```

Plot calibrated Ox data.
```{r}
oxygen %>% subset(experiment == "Mangrove_2" & light.dark == "Light") %>%
  ggplot(aes(OX_1, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(4,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = expression(O[2]~"("*mu*"mol L"^-1*")")) +
  #scale_x_continuous(limits = c(3,20)) +
  facet_wrap(~treatment) +
  theme_bw()

oxygen %>% subset(experiment == "Mangrove_2" & light.dark == "Dark") %>%
  ggplot(aes(OX_1, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(4,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = expression(O[2]~"("*mu*"mol L"^-1*")")) +
  #scale_x_continuous(limits = c(0,20)) +
  facet_wrap(~treatment) +
  theme_bw()
```

## 2. NO Profiles

```{r}
nitric_models <- nitric.cal.sum %>% 
  group_by(experiment, channel, light.dark, cal.ID) %>% 
  do(data.frame(., as.list(coef(lm(mean.mV~concentration, data = .))))) %>%
  rename_at(8:9, ~c("intercept", "slope"))

nitric <- nitric_models %>% subset(concentration == 0) %>% select(-mean.mV, -stdev.mV) %>%
  group_by(experiment, channel, light.dark) %>%
  pivot_wider(names_from = cal.ID, values_from = c(intercept, slope), names_vary = "slowest") %>%
  right_join(., nitric, by = join_by(experiment, channel, light.dark)) %>%
  subset(experiment == 'Mangrove_6') %>%
  select(experiment, channel, light.dark, incubation, treatment, 
         flume, core, replicate, analyte, time, depth, mV,
         intercept_1, slope_1, intercept_2, slope_2)

# calculate concentration for each calibration curve
nitric$NO_1 <- (nitric$mV - nitric$intercept_1) / nitric$slope_1
nitric$NO_2 <- (nitric$mV - nitric$intercept_2) / nitric$slope_2
```

Plot calibrated NO data.
```{r}
nitric %>% subset(light.dark == "Light") %>%
  ggplot(aes(NO_1, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  #scale_y_reverse(limits = c(8,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = expression("NO (nmol"~L^-1*")")) +
  #scale_x_continuous(limits = c(3,20)) +
  facet_wrap(~treatment) +
  theme_bw()

nitric %>% subset(light.dark == "Dark") %>%
  ggplot(aes(NO_2, depth/1000, color=as.factor(core))) +
  geom_point(aes(shape = as.factor(channel))) +
  geom_path(aes(shape = as.factor(channel))) +
  scale_y_reverse(limits = c(8,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = expression("NO (nmol"~L^-1*")")) +
  scale_x_continuous(limits = c(0,3000)) +
  facet_wrap(~treatment+channel, ncol = 2) +
  theme_bw()
```

## 3. N2O Profiles

```{r}
nitrous_models <- nitrous.cal.sum %>% 
  group_by(experiment, channel, light.dark, cal.ID) %>% 
  do(data.frame(., as.list(coef(lm(mean.mV~concentration, data = .))))) %>%
  rename_at(8:9, ~c("intercept", "slope"))

nitrous <- nitrous_models %>% subset(concentration == 0) %>% select(-mean.mV, -stdev.mV) %>%
  group_by(experiment, channel, light.dark) %>%
  pivot_wider(names_from = cal.ID, values_from = c(intercept, slope), names_vary = "slowest") %>%
  right_join(., nitrous, by = join_by(experiment, channel, light.dark)) %>%
  subset(experiment == 'Mangrove_2') %>%
  select(experiment, channel, light.dark, incubation, treatment, 
         flume, core, replicate, analyte, time, depth, mV,
         intercept_1, slope_1)

# calculate concentration for each calibration curve
nitrous$N2O_1 <- (nitrous$mV - nitrous$intercept_1) / nitrous$slope_1
```

Plot calibrated N2O data.
```{r}
nitrous %>% subset(light.dark == "Dark") %>%
  ggplot(aes(N2O_1, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(8,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(0,100)) +
  facet_wrap(~treatment) +
  theme_bw()

nitrous %>% subset(light.dark == "Light") %>%
  ggplot(aes(N2O_1, depth/1000, color=as.factor(core))) +
  geom_point() +
  geom_path() +
  scale_y_reverse(limits = c(8,-2)) +
  scale_color_discrete(name = "Core") +
  labs(y = "Depth (mm)", x = "Signal (mV)") +
  scale_x_continuous(limits = c(0,150)) +
  facet_wrap(~treatment) +
  theme_bw()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
