---
title: "Microprofile Experiments"
author: "Brett D. Jameson"
date: "2024-08-29"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyverse)
library(knitr)
library(plyr)
library(dplyr)
library(nplyr)
library(fuzzyjoin)
library(purrr)
library(broom)
```

## Mangrove Biogeochemistry (MBGC) Project - Microprofile Data Processing and Visualization

This R Markdown file contains code scripts associated with the processing and visualization of sediment microprofile data. 

Data were collected as part of the 2024 Mangrove Biogeochemistry project. Freshly collected mangrove sediment cores were incubated daily under varying nutrient amendment regimes and profiled to characterize vertical distributions of N2O, NO, H2S, and O2. Porewater profiles were obtained following light and dark incubations using Clark-type microelectrodes from Unisense. Temperatures were monitored throughout each experiment using HOBO temp/light pendant loggers.

## Metadata and HOBO logger data

Read and modify metadata.
```{r}
metadata <- read.csv(file="./MBGC.experiment.metadata.csv")
metadata <- metadata %>% mutate(start.date = as.POSIXct(start.date),
                                end.date = as.POSIXct(end.date),
                                low.tide = as.POSIXct(low.tide),
                                nuts.in = as.POSIXct(nuts.in),
                                dark.start = as.POSIXct(dark.start),
                                light.start = as.POSIXct(light.start)) %>%
                         select(!c(O2, N2O, NO, H2S, notes))
```

Read and modify HOBO logger files. 
```{r}
path.t = "./hobo-data/" #the location of HOBO files for all mangrove experiments

# List data files
file.names <- list.files(path = path.t, pattern = "csv$")  # list all csv file names in the folder
#file.names <- file.names[!grepl("metadata", file.names)]   # omit metadata from files to be read in as data

newnames <- c("obs", "date.time", "tempC", "coupler.attached", "host.connected", "stopped", "end.of.file") # create column names

fmt1 <- "%m/%d/%y %I:%M:%S %p"
df <- tibble(file.name = file.names) %>%
  mutate(experiment = gsub("^([^_]*_[^_]*)_.*$", "\\1", file.name)) %>% # Get experiment ID from filename
  mutate(flume = gsub("^[^_]+_[^_]+_[^_]+_[^_]+_([^_]+)_.*", "\\1", file.name)) %>% # Get flume ID from filename
  mutate(info = map(experiment, ~filter(metadata, experiment == .)),
         temp0 = map(file.name, ~read_csv(file.path(path.t, .), col_names = newnames, skip =2))) %>%
  nplyr::nest_mutate(temp0, date.time = as.POSIXct(date.time, format = fmt1))
```

Trim the HOBO logger data to remove data points outside the incubation interval.
```{r}
df_trimmed <- df %>%
  mutate(temp = map2(temp0, info, ~filter(.x, date.time >= .y$start.date & date.time <= .y$end.date))) 
```

Combine data files by experiment and add 'flume' identifier
```{r}
df_trimmed <- df_trimmed %>% select(!c(file.name, temp0)) %>% 
  group_by(experiment) %>%
  nest(temp = c(flume,temp)) %>%
   mutate(temp = map(temp, function(df) {
    # Unnest the 'temp' dataframe while keeping 'flume' intact
    df %>% 
       unnest_longer(temp) %>%
       unpack(temp)
  }))
```

# HOBO exploratory plots

Create function for generating exploratory temperature log plots
```{r pressure, echo=FALSE}
df_trimmed <- df_trimmed %>%
  mutate(temp_plot = pmap(list(.x = temp, .y = info, .z = experiment),
                          function(.x, .y, .z) {
                            ggplot(.x, aes(x = as.POSIXct(date.time), y = tempC, color = flume)) +
                            geom_vline(xintercept = .y$dark.start, color="darkblue") +
                            geom_vline(xintercept = .y$light.start, color="darkgreen") +
                            geom_vline(xintercept = .y$end.date, color = "darkred") +
                            geom_line() +
                            geom_point() +
                            scale_color_manual(values = c("#1B9E77", "#D95F02", "#7570B3", "#E6AB02")) +
                            labs(title = .z) +
                            theme_bw()
                          }
                        )
                      )
```

Pull out and inspect temperature logs for each experiment.
```{r}
df_trimmed %>% 
  filter(experiment == "Mangrove_5") %>%
  pull(temp_plot)
```

## Microprofile Data

Read and modify profile data.
```{r}
# Read data
profiles <- read.csv(file = "./compiled-data/MBGC.mangrove.microprofiles.2024.csv")
profiles <- profiles %>% dplyr::rename(date.time = time) %>%
                         mutate(date.time = as.POSIXct(date.time, format = "%Y-%m-%d %H:%M"),
                               date = as.Date(date, format = "%d/%m/%Y"),
                               flume = as.factor(flume),
                               core = as.factor(core),
                               treatment = as.factor(treatment),
                               replicate = as.factor(replicate),
                               channel = as.factor(channel))
```

Read and modify calibration data.
```{r}
cal.data <- read.csv(file = "./compiled-data/MBGC.sensor.calibrations.2024.csv")
cal.trim <- cal.data %>% group_by(experiment, analyte, light.dark, cal.ID, channel) %>%
                              distinct(experiment, .keep_all = TRUE) %>%
                              select(-c(cal.point, Calibration.Time.Point, concentration, signal.mV)) %>%
                         ungroup() %>% # Ungroup after processing
                         mutate(channel=as.factor(channel))
```

Next we will comined our profile and calibration data with the temperature logs and metadata, mapping by experiment ID.
```{r}
# Combine temperature logs with profiles and calibrations
df_combined <- df_trimmed %>%
  select(-temp_plot) %>%
  mutate(profiles = map(experiment, ~filter(profiles, experiment == .))) %>%
  mutate(profiles = map(profiles, as_tibble)) %>%
  mutate(cals = map(experiment, ~filter(cal.trim, experiment == .)))
```

Some of the cores contain two replicates when the first profile failed or looked suspicious. 
So, we need to select the best profile of the two replicates before downstream processing.
```{r}
df_combined <- df_combined %>%
  mutate(profiles = map(profiles, ~ {
    # Group by the variable within the nested dataframe
    .x %>%
      group_by(date, incubation, experiment, light.dark, treatment, flume, core) %>%
      # Use if_else to conditionally filter based on presence of replicate == 2
      filter(if (any(replicate == "2")) replicate == "2" else TRUE) %>%
      ungroup()  # Ungroup after processing
  }))
```

# Microprofile exploratory plots - raw data.

Create function for generating exploratory profile plots.
```{r pressure, echo=FALSE}
df_combined <- df_combined %>%
  mutate(profile_plot = map2(profiles, experiment, ~ggplot(.x, aes(x = mV, y = depth, color = core)) +
                                geom_path(aes(linetype = channel)) +
                                geom_point(aes(shape = channel)) +
                                labs(title = .y) +
                                scale_y_reverse() +
                                facet_wrap(~flume + light.dark + analyte, ncol = 4, scales = "free_x") +
                                theme_bw()
                             ))
```

Pull out and inspect profile raw (mV) data for each experiment
```{r}
df_combined %>% 
  filter(experiment == "Mangrove_6") %>%
  pull(profile_plot)
```

We can see from the Nitric Oxide experiments that we are getting sulfide interference on the sensor, as shown by the large spikes at the bottom of the profile domain. We want to remove these profiles before the drift correction procedure, as the correction relies on the assumption of zero analyte at the bottom of the profile. 

To accomplish this, we will tell R to look at the final mV reading for each profile and remove profiles where this value is significantly larger than the calibration intercept (zero-point.)

```{r}
# Step 1: Pre-process `cals` to create a lookup table with distinct intercepts per `cals_experiment`, `cals_light.dark`, and `cals_channel`
cals_lookup <- df_combined %>%
  unnest(cals, names_sep = "_") %>%  # Add suffix to avoid column name conflicts
  filter(cals_cal.ID == 1) %>%  # Keep only entries with cal.ID == 1
  select(cals_experiment, cals_light.dark, cals_channel, cals_analyte, cals_intercept) %>%  # Select relevant columns
  distinct(cals_experiment, cals_light.dark, cals_channel, .keep_all = TRUE)  # Keep distinct combinations of `light.dark` and `channel`

# Step 2: Apply `map2()` to `profiles` and map the intercept from `cals_lookup`
df_combined <- df_combined %>%
  mutate(profiles = map2(profiles, cals, ~ {
    # Add the relevant intercept value to each row in `profiles` based on `cals_experiment`, `cals_light.dark`, and `cals_channel`
    .x <- .x %>%
      left_join(cals_lookup, by = c("experiment" = "cals_experiment",
                                    "light.dark" = "cals_light.dark", 
                                    "channel" = "cals_channel"))  # Join with lookup to add intercept

    # Step 3: Apply filtering condition only if analyte is not "H2S"
    .x <- .x %>%
      group_by(experiment, date, incubation, light.dark, treatment, flume, core, analyte, channel) %>%
      filter(if (first(analyte) == "H2S") TRUE else !(last(mV) > (cals_intercept * 2))) %>%
      ungroup()
      
    # Step 4: Remove unnecessary columns
    .x %>%
      select(-cals_analyte, -experiment.y)
  }))
```

Redefine the plot function to ensure the correct profiles have been removed.
```{r pressure, echo=FALSE}
df_combined <- df_combined %>%
  mutate(profile_plot = map2(profiles, experiment, ~ggplot(.x, aes(x = mV, y = depth, color = core)) +
                                geom_path(aes(linetype = channel)) +
                                geom_point(aes(shape = channel)) +
                                labs(title = .y) +
                                scale_y_reverse() +
                                facet_wrap(~flume + light.dark + analyte, ncol = 4, scales = "free_x") +
                                theme_bw()
                             ))
```

Pull out and inspect profile raw (mV) data.
```{r}
df_combined %>% 
  filter(experiment == "Mangrove_6") %>%
  pull(profile_plot)
```

## Microprofile signal drift corrections
The profile data is currently in raw mV format. Before applying the calibrations to each experiment, we need to correct the sensor signal readout for changes in temperature during the profiling interval. 

We will start by merging the profile data with the corresponding HOBO temperature log data. We will then pull out the last three observations from each profile, assuming that these values represent 'true' zeros. These data points are then used to assess how the baseline signal changes with temperature.

```{r}
# Fuzzy join function to join by approximate date-time (within 5 minutes) and by flume
fuzzy_join_by_time <- function(df1, df2, tolerance = 5) {
  fuzzy_left_join(
    df1, df2,
    by = c("date.time" = "date.time", "flume" = "flume"),
    match_fun = list(
      function(x, y) abs(as.numeric(difftime(x, y, units = "mins"))) <= tolerance,
      `==`  # Exact match for flume
    )
  )
}

# Apply the fuzzy join to each profiles-temp pair, keeping only the closest temperature
df_combined <- df_combined %>%
  mutate(profiles = map2(profiles, temp, fuzzy_join_by_time)) %>%
  mutate(profiles = map(profiles, ~ {
    .x <- .x %>%
      # Calculate the time difference
      mutate(time_diff = abs(as.numeric(difftime(date.time.x, date.time.y, units = "mins")))) %>%
      # Group by the original date.time and flume to find the closest temperature
      group_by(date.time.x, flume.x) %>%
      # Keep only the row with the minimum time difference
      filter(time_diff == min(time_diff)) %>%
      ungroup() %>%
      select(-date.time.y, -time_diff) %>%  # Remove duplicate date.time from temp and the time_diff column
      dplyr::rename(date.time = date.time.x, flume = flume.x) %>%  # Rename columns
      select(-coupler.attached, -host.connected, -stopped, -end.of.file)  # Remove specified columns

    # Return the modified profiles
    .x
  }))

# Now create df_sliced using the modified profiles, grouping by both date.time and flume
df_sliced <- df_combined %>%
  mutate(zeros = map(profiles, ~
    .x %>%
      group_by(experiment, light.dark, treatment, flume, core, replicate, channel, analyte) %>%  # Include date.time in the grouping
      filter(if_else(analyte == "H2S", row_number() <= 3, row_number() > (n() - 3))) %>%
      ungroup()
  ))
```

# Plotting signal/temperature relationships.

Define a plot function for visualizing temp/mV relationships.
```{r}
df_sliced <- df_sliced %>%
  mutate(temp_correct = map2(zeros, experiment, ~ggplot(.x, aes(x = tempC, y = mV)) +
                                #geom_path(aes(linetype = channel)) +
                                geom_point(aes(shape = core, color = flume)) +
                                labs(title = .y) +
                                facet_wrap(~light.dark+channel, scales = "free", nrow = 2) +
                                theme_bw() +
                                facet_wrap(~channel+ light.dark, scales = "free")
                             ))
```

Pull out and inspect relationships by experiment/analyte.
```{r}
df_sliced %>% 
  filter(experiment == "Mangrove_8") %>%
  pull(temp_correct)
```

Try a drift correction based on the baseline signal versus time relationship.
```{r}
# Step 1: Define drift over time using linear regression
df_model <- df_sliced %>%
  mutate(drift_model = map(zeros, ~ .x %>%
    group_by(experiment, light.dark, channel, analyte) %>%
    # Convert date.time to numeric seconds from the minimum date.time in the group
    mutate(time_numeric = as.numeric(difftime(date.time, min(date.time), units = "secs"))) %>%
    do({
      model <- lm(mV ~ time_numeric, data = .)
      data.frame(
        slope = coef(model)[2],
        p_value = summary(model)$coefficients[2, "Pr(>|t|)"],
        slope_significant = summary(model)$coefficients[2, "Pr(>|t|)"] < 0.05
      )
    }) %>%
    ungroup()
  ))

# Step 2: Apply correction if slope is significant
df_corrected <- df_combined %>%
  mutate(profiles = map2(profiles, df_model$drift_model, ~ .x %>%
    left_join(.y %>% filter(slope_significant), by = c("experiment", "light.dark", "channel", "analyte")) %>%
    mutate(
      time_numeric = as.numeric(difftime(date.time, min(date.time), units = "secs")), # Apply same conversion
      corrected_mV = ifelse(slope_significant, mV - slope * time_numeric, mV)
    ) %>%
    select(-slope, -slope_significant, -time_numeric)
  ))
```

Now attempt to include temperature effects in the drift correction model.
```{r}
# Step 1: Define drift and temperature influence over time using a multiple regression model
df_model <- df_sliced %>%
  mutate(drift_model = map(zeros, ~ .x %>%
    group_by(experiment, light.dark, channel, analyte) %>%
    # Convert date.time to numeric seconds from the minimum date.time in the group
    mutate(time_numeric = as.numeric(difftime(date.time, min(date.time), units = "secs"))) %>%
    do({
      model <- lm(mV ~ time_numeric + tempC, data = .)  # Include tempC in the model
      data.frame(
        slope_time = coef(model)["time_numeric"],
        slope_tempC = coef(model)["tempC"],
        p_value_time = summary(model)$coefficients["time_numeric", "Pr(>|t|)"],
        p_value_tempC = summary(model)$coefficients["tempC", "Pr(>|t|)"],
        slope_time_significant = summary(model)$coefficients["time_numeric", "Pr(>|t|)"] < 0.05,
        slope_tempC_significant = summary(model)$coefficients["tempC", "Pr(>|t|)"] < 0.05
      )
    }) %>%
    ungroup()
  ))

# Step 2: Apply correction if either slope is significant
df_corrected <- df_combined %>%
  mutate(profiles = map2(profiles, df_model$drift_model, ~ .x %>%
    left_join(.y %>% filter(slope_time_significant | slope_tempC_significant), 
              by = c("experiment", "light.dark", "channel", "analyte")) %>%
    mutate(
      time_numeric = as.numeric(difftime(date.time, min(date.time), units = "secs")), # Apply same conversion
      corrected_mV = mV - 
                     ifelse(slope_time_significant, slope_time * time_numeric, 0) - 
                     ifelse(slope_tempC_significant, slope_tempC * tempC, 0)
    ) #%>%
#    select(-slope_time, -slope_tempC, -slope_time_significant, #-slope_tempC_significant, -time_numeric)
  ))
```

Create function for visualizing drift-corrected signal.
```{r pressure, echo=FALSE}
df_corrected <- df_corrected %>%
  mutate(profile_plot = map2(profiles, experiment, ~ggplot(.x, aes(x = corrected_mV, y = depth, color = core)) +
                                geom_path(aes(linetype = channel)) +
                                geom_point(aes(shape = channel)) +
                                labs(title = .y) +
                                scale_y_reverse() +
                                facet_wrap(~flume + light.dark + analyte, ncol = 4, scales = "free_x") +
                                theme_bw()
                             ))
```

Pull out and inspect profile raw (mV) data for each experiment
```{r}
df_corrected %>% 
  filter(experiment == "Mangrove_6") %>%
  pull(profile_plot)
```

## Applying sensor calibrations.

Now that the profiles have been corrected for signal drift, we can apply the calibration coefficients to calculate analyte concentrations.

```{r}
df_calibrated <- df_corrected %>%
  mutate(profiles = map2(profiles, cals, ~ {
    # Ensure 'channel' in .y is character to match .x
    .y <- .y %>% mutate(channel = as.character(channel))
    
    # Prepare calibration data with explicit column names for each calibration
    cal1 <- .y %>%
      filter(cal.ID == 1) %>%
      mutate(intercept1 = intercept, slope1 = slope) %>%
      select(experiment, light.dark, analyte, channel, intercept1, slope1)
    
    cal2 <- .y %>%
      filter(cal.ID == 2) %>%
      mutate(intercept2 = intercept, slope2 = slope) %>%
      select(experiment, light.dark, analyte, channel, intercept2, slope2)
    
    # Join calibration data to profiles
    .x <- .x %>%
      left_join(cal1, by = c("experiment", "light.dark", "analyte", "channel")) %>%
      left_join(cal2, by = c("experiment", "light.dark", "analyte", "channel"))
    
    # Calculate concentrations based on calibration parameters for each channel
    .x %>%
      mutate(
        concentration1 = if_else(!is.na(intercept1) & !is.na(slope1), 
                                 (corrected_mV - intercept1) / slope1, NA_real_),
        concentration2 = if_else(!is.na(intercept2) & !is.na(slope2), 
                                 (corrected_mV - intercept2) / slope2, NA_real_)
      ) %>%
      select(-intercept1, -slope1, -intercept2, -slope2)  # Clean up unnecessary columns
  }))
```


Plot the calibrated profiles. 
```{r pressure, echo=FALSE}
df_calibrated <- df_calibrated %>%
  mutate(profile_plot = map2(profiles, experiment, 
                             ~ggplot(.x, aes(x = concentration1, y = depth, color = core)) +
                                geom_path(aes(linetype = channel)) +
                                geom_point(aes(shape = channel)) +
                                labs(title = .y) +
                                scale_y_reverse() +
                                facet_wrap(~flume + light.dark + analyte, ncol = 4, scales = "free_x") +
                                theme_bw()
                             ))
```

Pull out and inspect profile raw (mV) data for each experiment
```{r}
df_calibrated %>% 
  filter(experiment == "Mangrove_8") %>%
  pull(profile_plot)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
